---
title: "Reproducible Research:Assignment_1"
author: "Vrinda Prabhu"
date: "Saturday 09 January 2016"
output: html_document
---

## Including the required library and reading the datasets required

The *raw_df* is a data frame constructed by reading directly from csv provided.It is then converted into data table format *raw_table*.
The *active_df* is data frame having removed all NA values.It is then converted into data table format *active_table*.
Also **date** in *active_df* is converted from factor to Date format.

```{r Loading library}
library(data.table)
library(ggplot2)

raw_df <- read.csv('activity.csv')
raw_table <- data.table(raw_df)

active_df <- raw_df[complete.cases(raw_df),]
active_df[,'date'] <- as.Date(active_df[,'date'], "%Y-%m-%d")
active_table <- data.table(active_df)

```

## REQUIREMENT 1

Here the total number of steps taken per day is calculated and a *histogram* and *barplot* plotted for the same.
[Difference between bar graphs and histograms can be found here](http://stattrek.com/statistics/charts/histogram.aspx?Tutorial=AP)
Owing to link shared,I was prompted for the bar graph.Please validate my understanding.

Also calculated is the mean and the median.

```{r Requirement 1,echo=FALSE}
setkey(active_table,date)
steps_per_day <- active_table[, list(steps=sum(steps)), by=key(active_table)] 

steps_per_day_plot_1 <- ggplot(steps_per_day, aes(steps))
steps_per_day_plot_1 <- steps_per_day_plot_1 + geom_histogram( binwidth = 1000,col="red", fill="blue" ,alpha = .2) +
                        xlab('Count of steps') +
                        ylab('Frequency') +
                        labs(title = 'Histogram of the total number of steps taken each day') +
                        theme(plot.title = element_text(hjust = 0.5))
steps_per_day_plot_1




steps_per_day_plot_2 <- ggplot(steps_per_day, aes(date,steps))
steps_per_day_plot_2 <- steps_per_day_plot_2 + geom_bar(stat = 'identity', binwidth = 1000,col="red", fill="blue" ,alpha = .2)+
                        xlab('Date') +
                        ylab('Count of steps') +
                        labs(title = 'Barplot of the total number of steps taken each day') +
                        theme(plot.title = element_text(hjust = 0.5))
steps_per_day_plot_2


steps_per_day_mean <- mean(steps_per_day$steps)
cat('Mean',steps_per_day_mean)

steps_per_day_median <- median(steps_per_day$steps)
cat('Median',steps_per_day_median)

```


## REQUIREMENT 2

Below code is to understand the average daily activity pattern.
This is achieved by a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.

Also the **_max_interval_** gives 5-minute interval, on average across all the days that contains the maximum number of steps.

```{r Requirement 2,echo=FALSE}
setkey(active_table,interval)
steps_per_interval <- active_table[, list(steps=mean(steps)), by=key(active_table)] 

steps_per_interval_plot <- ggplot(steps_per_interval, aes(interval,steps))
steps_per_interval_plot <- steps_per_interval_plot + geom_line(col="red", size=0.4) +
                           xlab('Interval') +
                           ylab('Number of steps') +
                           labs(title = 'Time series plot of the average number of steps taken') +
                           theme(plot.title = element_text(hjust = 0.5))
                          
steps_per_interval_plot

row_ind <- which.max(steps_per_interval$steps)
max_interval <- steps_per_interval[row_ind,interval]
cat('Interval with max number of steps',max_interval)

```

## REQUIREMENT 3
The **_na_number_** reports the total number of missing values in the dataset (i.e. the total number of rows with NAs)

NAs are filled using the mean number of steps for that 5-minute interval.This is achieved by finding mean steps across intervals and later imputing the same.Codewise the *data.table* chaining is used.

Histograms and bar plots of total number of steps taken per day are plotted with the imputed data.
Owing to my understanding,I also plotted the bar graph (feel its more informative).Please validate my understanding.

Also calculated is the mean and the median.

It is noticed that mean remains same whilst there is a slight change in the median.

```{r Requirement 3,echo=FALSE}
na_number <- table(complete.cases(raw_df))[1]
cat('Number of NAs',na_number)

raw_table <- raw_table[,steps := as.double(steps)
                      ][,steps.mean := mean(steps, na.rm = TRUE),by=interval  # mean across interval
                      ][is.na(steps), steps := steps.mean  # filling NAs with calculated mean
                      ][, steps.mean := NULL]  # dropping the column created
head(raw_table,10)

steps_per_day_wona <- raw_table[, list(steps=sum(steps)), by=date] 

steps_perday_wona_plot_1 <- ggplot(steps_per_day_wona, aes(steps))
steps_perday_wona_plot_1 <- steps_perday_wona_plot_1 + geom_histogram( binwidth = 1000,col="red", fill="blue" ,alpha = .2) +
                            xlab('Count of steps') +
                            ylab('Frequency') +
                            labs(title = 'Histogram of the total number of steps taken each day\n after missing values are imputed') +            
                            theme(plot.title = element_text(hjust = 0.5))
steps_perday_wona_plot_1


steps_perday_wona_plot_2 <- ggplot(steps_per_day_wona, aes(date,steps))
steps_perday_wona_plot_2  <- steps_perday_wona_plot_2  + geom_bar(stat = 'identity', binwidth = 1000,col="red", fill="blue" ,alpha = .2)+
                             xlab('Date') +
                             ylab('Count of steps') +
                             labs(title = 'Barplot of the total number of steps taken each day') +
                             theme(axis.text.x = element_text(angle = 90, hjust = 1))
steps_perday_wona_plot_2 


steps_per_day_wona_mean <- mean(steps_per_day_wona$steps)
cat('Mean of imputed data',steps_per_day_wona_mean)

steps_per_day_wona_median <- median(steps_per_day_wona$steps)
cat('Median of imputed data',steps_per_day_wona_median)

```


## REQUIREMENT 4

A new factor variable is created in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.This is done using the **decide_day** function,inherently using *weekdays()* function.

A panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days is later plotted.

```{r Requirement 4,echo=FALSE}
decide_day <- function(x){
  single_day <- weekdays(as.Date(x))
  ifelse(single_day %in% c('Saturday','Sunday'),'weekend','weekday')
}

raw_table <- raw_table[,'day' := decide_day(date),with = F]
head(raw_table)

setkey(raw_table,interval,day)
steps_per_interval_day <- raw_table[, list(steps=mean(steps)), by=key(raw_table)] 
steps_per_interval_day_plot <- ggplot(steps_per_interval_day,aes(interval,steps)) + geom_line(col='red') 
steps_per_interval_day_plot <- steps_per_interval_day_plot + facet_grid(day ~ .) +
                               xlab('Interval') +
                               ylab('Steps') +
                               labs(title = 'Panel plot comparing the average number of steps taken\n per 5-minute interval across weekdays and weekends') +
                               theme(plot.title = element_text(hjust = 0.5))
steps_per_interval_day_plot

```
